---
title: "M4: Outliers and Missing Data - Wind Speed Data"
author: "Luana Lima"
output: pdf_document
---


## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)

#New packages for M4
#install.packages("outliers")
library(outliers)
#install.packages("tidyverse")
library(tidyverse)

```

## Importing and preparing data

Today we will work with wind speed data. 

```{r}

#Importing time series data from text file#
getwd()
wind_data <- read.csv(file="../Data/Wind_Speed_PortArthurTX_Processed.csv",header=TRUE,stringsAsFactors = TRUE)

#creating data object
wind_data$DATE <- ym(wind_data$DATE)

```


## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. 

```{r}
ts_wind_data <- ts(wind_data$AWND.1,frequency=12, start = c(1984,1)) 

autoplot(ts_wind_data) +
  ylab("Average Wind Speed") 
#there are jumps in the graph

summary(wind_data$AWND.1) #shows 4 NA's
```


## Using pre-built function to clean series

tsclean() from package 'forecast' identify and replace outliers and missing values in a time series using linear interpolation. This should be used with some caution!! (as we might not want to replace the outliers just yet, i.e., before doing analysis and figuring out what type of outliers they are)

```{r}

clean_wind_data <- tsclean(ts_wind_data)  #object should be a time series

autoplot(clean_wind_data, series="Wind Clean") +
  autolayer(ts_wind_data, series="Wind Raw") +
  ylab("Average Wind Speed") 

#we should be seeing only 4 reds (= number of NA's).. but seeing more!


```

If you want to handle missing data, but do not want to remove outliers, here are two other options that will replace them using linear interpolation.

`forecast::na.interp()`
`zoo::na.approx()`

```{r, fig.width=15, fig.height=5}

#only replacing the missing values
#defining the package where we want to use the function from ::
clean_na_wind_data <- forecast::na.interp(ts_wind_data)

autoplot(clean_na_wind_data, series="Wind Clean (no NA)") +
  autolayer(clean_wind_data, series="Wind Clean (no NA nor outlier)") +
  #autolayer(ts_wind_data, series="Wind Raw") +
  ylab("Average Wind Speed")

```

```{r, fig.width=15, fig.height=5}
#only replacing the missing values
#defining the package where we want to use the function from ::
clean_na2_wind_data <- zoo::na.approx(ts_wind_data)

autoplot(clean_na2_wind_data, series="Wind Clean (no NA forecast)") +
  autolayer(clean_na_wind_data, series="Wind Clean (no NA zoo)") +
  #autolayer(clean_wind_data, series="Wind Clean (no NA nor outlier)") +
  #autolayer(ts_wind_data, series="Wind Raw") +
  ylab("Average Wind Speed")

```



